Natural Selection in the 20th Century

### 2020-11-13 why-negative-selection.Rmd
Rewrite discussion of economic theory. Add some more papers.

---
title: "Natural Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "`r Sys.Date()`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. The strength of natural selection on a polygenic score can be predicted by its correlation with earnings and education. Effects are larger among low-income and less educated individuals, younger parents, and people not living with a partner. Effect sizes are larger when we correct for ascertainment bias. The direction of natural selection is reversed among older parents, or  controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings can either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
---

@@

* Happel et al. 1994
   - childbirth is chosen to minimize loss of earnings from interrupted work.
   - early childbirth if your skills decay a lot  while childrearing. (No point
     gaining them then losing them.) Late childbirth otherwise.
   - late childbirth if you have acquired skills before marriage.
   - Capital-market imperfections make it wise to delay childbirth (till man has
     high income to cope while wife is at home). More so for men's wages which
     rise faster.
   - empirically: (among married couples) 
     - high-skill occupation women have children later. Higher-earning men have 
     them earlier.

@@

Why do selection effects have *opposite* signs across older and younger mothers
at first live birth, and opposite signs when AFLB is controlled for? A possible
answer is given by the economic theory of fertility [@becker1960economic; @willis1973new].
According to this, increases in potential earned income affect fertility via two
opposing channels. There is an "income effect" by which children become more
affordable, like any other good. There is also a "substitution effect": since
childrearing has a cost in time, the opportunity cost of childrearing increases
if one's market wage is higher. The income effect would lead higher earners to
have more children. The substitution effect would lead higher earners to have
fewer children. It is often assumed that the substitution effect will dominate
for lone parents, or those in unstable relationships, since they have less
opportunity to share childcare responsibilities, while the income effect will
dominate for couples who are able to reap the gains from specialization.

If so, then genetic characteristics which affect one's earnings
potential in the labour market may lead to opposing effects on
fertility. Genetic variants which improve one's earnings may increase
fertility among couples, but decrease fertility among single parents.
They may also decrease the likelihood of becoming a young mother, if
this is correlated with single motherhood. Lastly, if lifetime number of
sexual partners predicts single parent versus two-parent status, the
theory could also explain the different effect sizes across people with
more and fewer sexual partners.

This explanation predicts that the strength of polygenic scores' effects
on fertility will correlate with the strength of their relationship with
earnings. To test this, we calculate the correlation of each PGS with
estimated earnings from respondent's first job[^earnings-est], and with education
level, a phenotype which predicts earnings. We then relate these
measurements to effect sizes on number of children. There is a clear
negative relationship between a score's effect on fertility, and its
correlation with earnings or education. If we control for age at first live
birth, the relationship changes sign and becomes positive  (Figure
\@ref(fig:plot-cors-earnings-educ)).

@@

```{r plot-cors-earnings-educ, fig.cap="Effect sizes on number of children, with and without controls for age at first live birth, by correlations with earnings and education. Each point represents one PGS."}

@@

cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
# TODO: use age_left_edu when this includes college
cors_educ <-   cor(famhist[score_names], famhist$edu_qual, use = "pairwise")

@@

dfr %>% 
      mutate(
        Controls = fct_relevel(Controls, "None")
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        facet_grid(Correlation ~ Controls, labeller = label_both) + 
        my_vline + 
        my_hline +
        labs(x = "Effect on number of children", y = "Correlation") +
        scale_colour_brewer(type = "qual", palette = 2) +
        theme(legend.position = "none")

@@

@@

Research on natural selection is controversial. It brings to mind eugenics,
which has been used to justify mass murder and human rights violations, in the
past and possibly the present [@mao1998chinese; @sleeboom2011genetic;
@zenz2020]. Whilst we claim scientific legitimacy for our empirical results, we
have no special authority to speak on how society should react to
them. This can only be resolved by open debate, in which scientists take part as
equals among many. We therefore choose not to discuss policy implications here.

@@

## Genetic correlations with EA3

@@

## Number of children

>>>>

---
title: "Natural Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "`r Sys.Date()`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. The strength of natural selection on a polygenic score can be predicted by its correlation with earnings and education. Effects are larger among low-income and less educated individuals, younger parents, and people not living with a partner. Effect sizes are larger when we correct for ascertainment bias. The direction of natural selection is reversed among older parents, or  controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
---

@@

@@

* Happel et al. 1984
   - childbirth is chosen to minimize loss of earnings from interrupted work.
   - early childbirth if your skills decay a lot  while childrearing. (No point
     gaining them then losing them.) Late childbirth otherwise.
   - late childbirth if you have acquired skills before marriage.
   - Capital-market imperfections make it wise to delay childbirth (till man has
     high income to cope while wife is at home). More so for men's wages which
     rise faster.
   - empirically: (among married couples) 
     - high-skill occupation women have children later. Higher-earning men have 
     them earlier.

- Baudin et al. 2015
  * look at childlessness vs. fertility (extensive/intensive margin)
  * some people are too poor to afford children
  * 'a husband alleviates the time cost of having children'. But for some groups (see appendix)
    there is a +ve income/fertility gradient
  * fertility decreases with female educn due to subst effect
  * baby boom and bust in US: a decline in childlessness caused by poverty, then an increase
    caused by substn effect ('opportunity cost')
  * fertility decreases with educn both for single mothers and married women
    - slightly (25%) higher decrease per year of educn (Table 2) for single mothers

Why do selection effects have *opposite* signs across older and younger mothers
at first live birth, and opposite signs when AFLB is controlled for? A possible
answer is given by the economic theory of fertility 
[@becker1960economic; @willis1973new; @becker1976child].
According to this, increases in potential earned income affect fertility via two
opposing channels. There is an "income effect" by which children become more
affordable, like any other good. There is also a "substitution effect": since
childrearing has a cost in time, the opportunity cost of childrearing increases
if one's market wage is higher. The income effect would lead higher earners to
have more children. The substitution effect would lead higher earners to have
fewer children. If so, then genetic variants which affect earnings
potential in the labour market may cause opposing effects on
fertility. The income effect will cause natural selection in favour of 
earnings-increasing variants. The substitution effect will do the reverse.

This theory predicts that genetic variants will be selected for (or against) in
proportion to their effect on earnings. Figure
\@ref(fig:plot-cors-earnings-educ) plots selection effects on each PGS against
that PGS's correlation with earnings in a respondent's first job[^earnings-est],
and against its correlation with educational attainment, a predictor of lifetime
earnings. The raw relationships (left column) are strongly negative. If we plot
selection effects controlling for age at first live birth (right column), the
effect reverses and becomes positive. This is evidence that the labour market
plays an important role in natural selection, with substitution effects
dominating overall.

@@

```{r plot-cors-earnings-educ, fig.cap="Selection effects, with and without controls for age at first live birth, by correlations with earnings and educational attainment. Each point represents one PGS."}

@@

cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
# TODO: use age_left_edu when this includes college
# TODO:
cors_educ <-   cor(famhist[score_names], famhist$age_fulltime_edu, use = "pairwise")

@@

dfr %>% 
      mutate(
        Controls = fct_relevel(Controls, "None")
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        facet_grid(Correlation ~ Controls, labeller = label_both) + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "Correlation") +
        scale_colour_brewer(type = "qual", palette = 2) +
        theme(legend.position = "none")

@@

Suppose that people are aware of their endowments of human capital, and 
that education and innate human capital are complements. They choose how long
to stay in education. After leaving education they enter the labour market and
form families. If so, then people with less human capital will leave education 
earlier and have their first child earlier. Whilst in the labour market, higher
earners will have more children (i.e. income effects dominate). These 
assumptions will lead to the pattern we observe where earnings-increasing
variants lead to later first births (and fewer children overall), but more
children for any given AFLB.

The theory can also explain why natural selection is weaker among higher-income
parents. @becker1976child show that if parents care about child quality as well
as quantity, under certain assumptions the income effect will be stronger at
higher income levels; this can lead to a U-shaped relationship between income
and fertility.

Lastly, it is often assumed that the substitution effect will dominate
for lone parents, or those in unstable relationships, since they have less
opportunity to share childcare responsibilities, while the income effect will
dominate for couples who are able to reap the gains from specialization. If so,
then earnings-increasing genetic variants may decrease fertility more among single 
parents than among couples, as in our data. The same idea could explain our
results for lifetime number of sexual partners, if this is associated with
relationship stability.

Empirical work shows similar relationships between *earnings* or *education* and
fertility, to the relationships shown here between *genetics* and fertility.
Higher income is associated with lower fertility [@becker1960economic;
@jones2006economic].
@cohen2013financial find that income decreases fertility at low income levels
but increases it at higher income levels. In a US sample, fertility decreases
with education faster among single mothers than married mothers 
[@baudin2015fertility].

```{r calc-res-earn-educ}

drake::loadd(res_ee_control)
drake::loadd(res_all)

res_ee_control %<>% filter(term == score_name)
n_05 <- sum(res_ee_control$p.value < 0.05)
n_05_33 <- sum(res_ee_control$p.value < 0.05/33)

The economic theory predicts that natural selection on polygenic scores is mediatet
by labour market earnings. As a final test of this,
we re-estimate equation (\@ref(eq:regression)) for each PGS, 
controlling for earnings in first job and education levels. Effect sizes are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only `r n_05_33` 
out of 33 PGS are significant at $p = 0.05/33$.

@@

Research on natural selection is controversial. It brings to mind eugenics,
which has been used to justify mass murder and human rights violations, in the
past and possibly the present [@mao1998chinese; @sleeboom2011genetic;
@zenz2020]. Whilst we claim scientific legitimacy for our empirical results, we
have no special authority to speak on how society should react to
them. This is a matter for open debate, in which scientists take part as
equals among many. We therefore choose not to discuss policy implications here.

@@

\FloatBarrier
## Controlling for earnings and education

```{r plot-res-earn-educ, fig.cap = "Selection effects controlling for earnings in first job (estimated by median earnings from ASHE 2007 for the SOC2000 job code) and education (left education before 16, 16-18, or after 18). Raw effects are shown for comparison. Lines are 95% confidence intervals uncorrected for multiple testing."}

drake::loadd(res_ee_control)
drake::loadd(res_all)

res_ee_control %<>% filter(term == score_name)

res_all %<>% 
  filter(term == score_name, reg.type == "raw", dep.var == "N children") %>% 
  select(-dep.var, -reg.type)

res_compare <- bind_rows(
                 Raw = res_all, 
                 Controlled = res_ee_control, 
                 .id = "Regression"
               )

standard_ggplot(res_compare, fill_col = Regression, n_regs = 33) +
  geom_linerange(
    aes(
      xmin = estimate - 2 * std.error, 
      xmax = estimate + 2 * std.error
    ), 
    alpha = .5
  )

\FloatBarrier
## Genetic correlations with EA3

@@

\FloatBarrier
## Number of children
